---
title: "Data Preparation"
format: html
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true
---

This document will go through how we prepare the data that will be used for the Shiny application.

# 1 Setup

## 1.1 Loading Packages

We will use the following packages to prepare the data.

-   [jsonlite](https://cran.r-project.org/web/packages/jsonlite/index.html) - To parse JSON

-   [knitr](https://cran.r-project.org/web/packages/knitr/index.html) - For better table displays

-   [tidyverse](https://www.tidyverse.org/) - Data science tools

-   [tidygraph](https://cran.r-project.org/web/packages/tidygraph/index.html) - For graph manipulations

-   [igraph](https://cran.r-project.org/web/packages/igraph/index.html) - Contains functions for network analysis

```{r}
pacman::p_load(jsonlite, knitr, tidyverse, tidygraph, igraph)
```

## 1.2 Loading Data

We will load the provided [VAST Mini-Challenge 3](https://vast-challenge.github.io/2024/MC3.html) dataset, a json file.

```{r}
mc3_data <- fromJSON("data/mc3.json")
glimpse(mc3_data)
```

There are **60,520 nodes** and **75,817 edges** in the data.

# 2 Graph Elements

## 2.1 Extracting nodes and edges

We will first extract the nodes and edges.

::: panel-tabset
### Nodes

```{r}
mc3_nodes_raw <- as_tibble(mc3_data$nodes)
glimpse(mc3_nodes_raw)
```

We will only retain the following columns:

-   **id**: to serve as the identifier for the node
-   **type**: to differentiate people from companies in the graph.
-   **ProductServices**: to identify the products of services a business offer

```{r}
mc3_nodes_lite <- mc3_nodes_raw %>%
  select(id, type, ProductServices) %>%
  rename(product_services = ProductServices)
```

### Edges

```{r}
mc3_edges_raw <- as_tibble(mc3_data$links)
glimpse(mc3_edges_raw)
```

We will only retain the following columns:

-   **source**: to identify the actor of the relationship, corresponds to id in nodes.
-   **target**: to identify the receiver of the relationship, corresponds to id in nodes.
-   **type**: to identify the type of the relationship
-   **start_date**: to identify when the relationship started
-   **end_date**: to identify when the relationship ended

```{r}
mc3_edges_lite <- mc3_edges_raw %>% select(source, target, type, start_date, end_date)
```
:::

## 2.2 Deeper look at *type*

Both the `nodes` and `edges` have `type` which contains the type of the nodes and edges. We will assign a `supertype` and a `subtype` from `type`.

::: panel-tabset
### Nodes

```{r}
mc3_nodes_lite %>%
  group_by(type) %>%
  summarize(count = n()) %>%
  arrange(-count) %>%
  kable()
```

`supertype` - type of entity, either *Person* or *Organization*

`subtype` - subcategory of *supertype*, e.g., *Company, FishingCompany, CEO*

### Edges

```{r}
mc3_edges_lite %>%
  group_by(type) %>%
  summarize(count = n()) %>%
  arrange(-count) %>%
  kable()
```

`supertype` - type of relationship, either *Ownership*, *Employment, Relationship.*

`subtype` - subcategory of *supertype*, e.g., *Shareholdership, BeneficialOwnership, FamilyRelationship*
:::

## 2.3 Dates

Consider the date fields, e.g. `start_date`.

```{r}
mc3_edges_lite %>% select(start_date) %>% glimpse()
```

It is using the [ISO 8601](https://en.wikipedia.org/wiki/ISO_8601) format that includes the time component. We are only interested in the date component so we will just get the **first 10 characters**.

```{r}
substr("2016-10-29T00:00:00", 1, 10)
```

## 2.4 Node Aliases

As the nodes have long names, using them as labels in the visualization is not the best way as the text will cover important information.

We will generate them from the first character of each word. We will define a function to provide this capability.

```{r}
to_initials <- function(name) {
  strsplit(name, "[^A-Za-z0-9']+")[[1]] %>%  # Split when non-alphanumeric
    substr(1, 1) %>% # Get first letter
    paste0(collapse = "") %>%
    substr(1, 4) # Get first 4 letters only as some names are still too long
}
```

```{r}
to_initials("SouthSeafood Express Corp")
```

## 2.5 Other considerations

::: panel-tabset
### Filtering

We will add an **included** column to the nodes and edges for filtering purposes so that we can show or hide them depending on the filtering criteria.

This is particularly useful in network visualization.

### Graph attributes

We must also rename the columns for compatibility with *igraph* and *tidygraph*.

For edges, we will rename `source` and `target` to `from` and `to` respectively.

For nodes, we will rename `id` to `name`.
:::

# 3 Preparing the Nodes

## 3.1 Shaping the data

We will now prepare the nodes according to the above considerations.

```{r}
mc3_nodes_clean <- mc3_nodes_lite %>%
  rename(name = id) %>%
  mutate(
    alias = sapply(name, to_initials),
    supertype = strsplit(type, ".", fixed=TRUE) %>% sapply('[', 2),
    # Get the last type as subtype. In the case of Entity.Person,
    # both supertype and subtype are "Person".
    subtype = strsplit(type, ".", fixed=TRUE) %>% sapply(tail, n=1),
    included = 1
  ) %>% select(name, alias, supertype, subtype, product_services, included)
```

## 3.2 Checking the type fields

Let's confirm if the types have been mapped correctly to the corresponding `supertype` and `subtype`.

```{r}
mc3_nodes_clean %>%
  group_by(supertype, subtype) %>%
  summarize(count = n()) %>%
  arrange(-count) %>%
  kable()
```

## 3.3 Checking the rest of the data

Let's also inspect the rest of the data if they are in the form we need.

```{r}
mc3_nodes_clean %>% head() %>% kable()
```

The alias was successfully generated based on the node name. The dataframe also has all the columns we need.

# 4 Preparing the Edges

## 4.1 Shaping the data

With the previous considerations, we will shape the edge data.

```{r}
mc3_edges_clean <- mc3_edges_lite %>%
  rename(from = source, to = target, ) %>%
  mutate(
    supertype = ifelse(
      grepl("Event.Owns", type),
      "Ownership",
      ifelse(grepl("Relationship", type), "Relationship", "Employment")
    ),
    subtype = strsplit(type, ".", fixed = TRUE) %>% sapply(tail, n = 1),
    # Convert date strings to datetime
    start_date = substr(start_date, 1, 10) %>% as_date(),
    end_date = substr(end_date, 1, 10) %>% as_date(),
    included = 1
  ) %>%
  filter(from != to) %>%
  group_by(from, to, supertype, subtype, start_date, end_date, included) %>%
  summarize(weight = n())
```

## 4.2 Checking the type fields

Let's confirm if the types have been mapped correctly to the corresponding `supertype` and `subtype`.

```{r}
mc3_edges_clean %>%
  group_by(supertype, subtype) %>%
  summarize(count = n()) %>%
  arrange(-count) %>%
  kable()
```

## 4.3 Checking the rest of the data

Let's also inspect the rest of the data if they are in the form we need.

```{r}
mc3_edges_clean %>% head() %>% kable()
```

The dates columns only have the date components, not the time. The dataframe also has all the columns we need.

# 5 Preparing the Networks

## 5.1 Supernetwork

We will generate supernetwork containing all the nodes and edges we prepared.

```{r}
supernetwork = tbl_graph(
  edges = mc3_edges_clean,
  nodes = mc3_nodes_clean,
  directed = TRUE
)
```

Let's check the size of this network.

::: panel-tabset
### Node Count

```{r}
vcount(supernetwork)
```

### Edge Count

```{r}
ecount(supernetwork)
```
:::

We will save this as an RDS file for use in the Shiny app.

```{r}
write_rds(supernetwork, "data/rds/supernetwork.rds")
```

## 5.2 Filter by subnetwork

The supernetwork is very large and not suitable for visualization as it requires a lot of computing resources to visualize.

Hence, we will create a function that will enable us to focus on the network of a given node. We will define a function `extract_subnetwork` for this.

```{r}
extract_subnetwork <- function(graph, node_name, distance=-1) {
  # negative distance will show full graph
  node <- which(V(graph)$name == node_name)
  
  if(length(node) == 0) {
    # Return empty graph
    return(tbl_graph())
  }
  
  distance <- ifelse(distance < 0, length(graph), distance)
  vertices <- ego(graph, nodes = node, order = distance)[[1]]
  igraph_subgraph <- induced_subgraph(graph, vids = vertices)
  nodes_df <- as_data_frame(igraph_subgraph, what = "vertices")
  edges_sf <- as_data_frame(igraph_subgraph, what = "edges")
  tbl_graph(nodes=nodes_df, edges=edges_sf, directed=is_directed(graph))
}
```

::: {.callout-tip appearance="simple"}
### About the function

This function generates a subnetwork from a **graph** based on the nodes in proximity to **reference node**.

The size of the network can be controlled by the **distance** of the other nodes from the reference node. If the distance is negative, it will include all nodes connected in any way to the reference node.

It uses `ego` from `igraph` to figure out which nodes are connected within a given distance from a node.
:::

## 5.3 Filter by date

Next, to enable inspecting temporal patterns, we will filter the edges and nodes based on their existence on the given date.

We will define `extract_network_snapshot` to enable this filtering.

```{r}
extract_network_snapshot <- function(graph, datestring, delete = FALSE) {
  date <- as_date(datestring)
  
  graph_nodes = as_data_frame(graph, what = "vertices")
  graph_edges = as_data_frame(graph, what = "edges")
  
  if(is.na(date) || vcount(graph) == 0) {
    return(graph)
  }
  
  # Assume transition is at 12 AM of given date
  graph_edges <- graph_edges %>%
    mutate(
      included = ifelse(is.na(start_date) | (
        start_date <= date &
          (is.na(end_date) |
             end_date > date)
      ), 1, 0)
    )
  
  filtered_edges <- graph_edges %>% filter(included == 1)
    
  graph_nodes <- graph_nodes %>%
    mutate(included = (
      name %in% filtered_edges$from | name %in% filtered_edges$to
    ))
  
  if(!delete) {
    return(
      tbl_graph(nodes = graph_nodes,
                edges = graph_edges,
                directed = is_directed(graph))
    )
  }
  
  tbl_graph(nodes = graph_nodes %>% filter(included == 1),
            edges = filtered_edges,
            directed = is_directed(graph))
}
```

::: {.callout-tip appearance="simple"}
### About the function

This function sets `included` to `true` if the edge is active during the given date. For the nodes, they are considered active if they are connected to at least one other node at that point in time.

There is also an option to delete the inactive elements altogether, which is useful to calculate measures of centrality on the network structure at that point in time.
:::

# 6 Preparing the power graph

To figure out who are the most influential entities within the network, we have to look at the power dynamics in each relationship. This entails generating a graph from the relationship graph that reflects these dynamics.

## 6.1 Arrow direction

As we want to investigate the flow of power and resources, the direction will be from the **less powerful to more powerful** entity.

## 6.2 Weights

For the purpose of our analysis, we will score each relationship flow according to the following:

-   Benefits from the resources: 1 (+1 if they are the owner of a company)

-   A decision-maker on where resources go: 2 (+1 if they are the ultimate decision maker)

-   Influential on the other person: 1 (only if target is a person)

We will use the total as the weights for each edge. Lastly, we will **rename** the relationship to prevent confusion in case of direction reversal.

## 6.3 Score table

::: panel-tabset
### BeneficialOwner

Beneficial owners benefit from the resources of the company.

While BeneficialOwners may or may not be Shareholders in the same company, they hold the ultimate decision-making power.

|                            |         |
|----------------------------|---------|
| Current **From**           | Owner   |
| Current **To**             | Company |
| New **From** (Lower power) | Company |
| New **To** (Higher power)  | Owner   |
| New **subtype**            | OwnedBy |
| Beneficiary                | 1 + 1   |
| Decision-maker             | 2 + 1   |
| Influential (on person)    | 0       |
| **Total**                  | **5**   |

: Beneficial Ownership {.bordered .striped}

### Shareholdership

Shareholders may receive proceeds from the profits of the company.

They have some decision-making power within the company. However, it is the Beneficial Owner that ultimately gets the final say in decisions.

|                            |                |
|----------------------------|----------------|
| Current **From**           | Shareholder    |
| Current **To**             | Company        |
| New **From** (Lower power) | Company        |
| New **To** (Higher power)  | Shareholder    |
| New **subtype**            | HasShareholder |
| Beneficiary                | 1              |
| Decision-maker             | 2              |
| Influential (on person)    | 0              |
| **Total**                  | **3**          |

: Shareholdership {.bordered .striped}

### WorksFor

Employees are contractually obligated to act in the interests of their employer.

Employers also influence what their employees do in the business.

|                            |         |
|----------------------------|---------|
| Current **From**           | Owner   |
| Current **To**             | Company |
| New **From** (Lower power) | Company |
| New **To** (Higher power)  | Owner   |
| New **subtype**            | OwnedBy |
| Beneficiary                |         |
| Decision-maker             | 2       |
| Influential (on person)    | 1       |
| **Total**                  | 3       |

: WorksFor {.bordered .striped}

### FamilyRelationship

As it is impossible to figure out the power dynamics within a family, we will retain direction and assume that the arrow points to the person that benefits more from the relationship.

|                            |                    |
|----------------------------|--------------------|
| Current **From**           | Person 1           |
| Current **To**             | Person 2           |
| New **From** (Lower power) | Person 1           |
| New **To** (Higher power)  | Person 2           |
| New **subtype**            | FamilyRelationship |
| Beneficiary                | 1                  |
| Decision-maker             | 0                  |
| Influential (on person)    | 1                  |
| **Total**                  | **2**              |

: FamilyRelationship {.bordered .striped}
:::

::: callout-note
### Model Revision

From the results of visualization in the original [Take-home Exercise 3](https://isss608-kjcpaas.netlify.app/take-home_exs/ex3/take-home_ex3#table-1), the model for the power graph resulted in the influence `FamilyRelationship` being too strong that it overrode all power held from business activities.

We revised this to be unidirectional to lessen the impact and we cannot assume that family members have the same sharing dynamic.
:::

## 6.4 Converting relationship edges to power flow

Using the table above, we will create a function to convert relationship edges to power flow. We will define a function `convert_edges_to_power_flow` to take care of converting edges to power graph

```{r}
#| code-fold: true
#| code-summary: "Show the code for convert_edges_to_power_flow()"
convert_edges_to_power_flow <- function(edges) {
  # Employee -> Employer, weight: 3
  works_for <- edges %>% filter(subtype == "WorksFor") %>% mutate(weight = 3)
  
  # Person1 <-> Person2, weight: 2
  family <- edges %>% filter(subtype == "FamilyRelationship") %>% mutate(weight = 2)
  
  # Shareholder <- Company, weight: 3
  shareholder <- edges %>% filter(subtype == "Shareholdership") %>%
    mutate(
      temp = from,
      from = to,
      to = temp,
      weight = 3,
      # Rename to prevent confusion due to reversed arrows
      subtype = "HasShareholder",
    ) %>%
    select(from, to, supertype, subtype, start_date, end_date, weight, included)
  
  # BeneficialOwner <- Company, weight: 5
  owner <- edges %>% filter(subtype == "BeneficialOwnership") %>%
    mutate(
      temp = from,
      from = to,
      to = temp,
      weight = 5,
      # Rename to prevent confusion due to reversed arrows
      subtype = "OwnedBy",
    ) %>%
    select(from, to, supertype, subtype, start_date, end_date, weight, included)
  
  works_for %>%
    rbind(family) %>%
    rbind(shareholder) %>%
    rbind(owner)
}
```

Let us check the resulting edges table to see if each edge type has been mapped correctly.

::: panel-tabset
### Relationship

```{r}
supernetwork %>%
  as_data_frame(what = "edges") %>%
  group_by(supertype, subtype) %>%
  summarize(count = n()) %>%
  arrange(-count) %>%
  kable()
```

### Power

```{r}
supernetwork %>%
  as_data_frame(what = "edges") %>%
  convert_edges_to_power_flow() %>%
  group_by(supertype, subtype) %>%
  summarize(count = n()) %>%
  arrange(-count) %>%
  kable()
```
:::
